---
title: "Predicting cash flows using generalized additive models"
author: "Daniel Saxton"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(gam)
```

## Outline

- Problem description
- Probabilistic model
- Generalized additive models
- R code

## Problem

Given information on a customer and the terms of a loan, determine how much cash we expect to collect from this customer.

- Loan amount
- Loan term
- Payment amount
- Covariate values

## Two key behaviors

- Default
- Prepayment

Both of these are highly dependent on where a customer is in the loan life cycle.

## Probabilistic model

Associate with each loan two independent processes

- $T_d$: time until default
- $T_p$: time until prepayment

## Probabilistic model

Whether a loan with term $n$ is defaulted, prepaid or paid at term corresponds to these events

- $T_d > n$, $T_p > n$: loan is paid at term
- $T_d \leq n$, $T_d \leq T_p$: loan is defaulted
- $T_p \leq n$, $T_p < T_d$: loan is prepaid

In words, which of $T_d$, $T_p$ or $n$ is the smallest determines the outcome of the loan, with $T_d$ winning ties.

## Expected cash

If $X$ is the total collected on a given loan and $Y_i$ the amount collected at installment $i$

$$
\text{E}(X) = \sum_{i=1}^{n} \text{E}(Y_i) .
$$

## Expected cash

Also if $c$ is the regularly-scheduled payment amount and $c_i$ the payment amount plus remaining principal at time $i$ then

$$
\text{E}(Y_i) = c P(T_d > i) P(T_p > i) + c_i P(T_d > i) P(T_p = i) .
$$

## Hazard function

For a positive integer-valued event time $T$ define the hazard function for $k \in \mathbb{N}_+$

$$
h(k) = P(T = k \mid T \geq k) .
$$

We also have the nice relationship

\begin{align}
P(T > k) &= \prod_{j=1}^{k} \left [ 1 - h(j) \right ] .
\end{align}

## Hazard function

- The hazard function is a probability and so can be modeled using any probabilistic classifier

- The key is to properly format the data and treat the argument of the hazard function as a covariate

## Sample data

```{r, echo=FALSE}
data = data.frame(event = c('default', 'prepay', 'default', 'full_pay'),
                  install_num = c(1, 3, 2, 4),
                  term = c(5, 15, 3, 4),
                  payment = c(300, 200, 200, 300),
                  amount = c(1000, 2000, 500, 1500),
                  x = rpois(4, lambda=6))
data
```

After formatting:

```{r, echo=FALSE}
expanded = data[c(rep(1, 1), rep(2, 3), rep(3, 2), rep(4, 4)),]
expanded$install_num = c(1:1, 1:3, 1:2, 1:4)
expanded$default = c(1, rep(0, 4), 1, rep(0, 4))
expanded$prepay = c(rep(0, 3), 1, rep(0, 6))
expanded$event = NULL
rownames(expanded) = NULL
expanded
```

## Generalized additive models

Let $\mu_i \in \Theta$ be the mean of $y_i$, $g : \Theta \mapsto \mathbb{R}$ a link function and $\{ x_{ij} \}_{j=1}^{p}$ a set of covariate values.

$$
g(\mu_i) = s_0 + \sum_{j=1}^{p} s(x_{ij})
$$

where the $\{ s_j \}_{j=1}^{p}$ are smooth functions and $s_0$ a constant.  

## Advantages

- Less restrictive than GLMs, can model non-linearities
- Easy to interpret
- Useful as an exploratory tool, e.g., finding transformations

## Disadvantages

- More prone to overfit compared to linear models
- Implementation is more challenging

## Additive models in R

- `gam` (Trevor Hastie)
- `mgcv` (Simon Wood)

## Example

$$
y_i = \log(x_i) + \epsilon_i
$$

```{r, fig.align='center'}
x = rexp(100)
y = log(x) + rnorm(100)
data_set = data.frame(x, y)
plot(data_set$x, data_set$y, xlab='x', ylab='y')
```

## Smooth fit

```{r, echo=TRUE, message=FALSE, fig.align='center'}
model = gam(y ~ s(x), data=data_set)
plot(model)
```

## Smooth fit

```{r, echo=TRUE, message=FALSE, fig.align='center'}
model = gam(y ~ s(x, df=2), data=data_set)
plot(model)
```

## Example

$$
y_i = x_i^2 + \epsilon_i
$$

```{r, fig.align='center'}
x = rexp(100)
y = x^2 + rnorm(100)
data_set = data.frame(x, y)
plot(data_set$x, data_set$y, xlab='x', ylab='y')
```

## Smooth fit

```{r, echo=TRUE, message=FALSE, fig.align='center'}
model = gam(y ~ s(x), data=data_set)
plot(model)
```

## Additive logistic regression

For data of the form $\{y_i, x_{i1}, \ldots , x_{ip} \}_{i=1}^{n}$ where $y_i \sim$ Bernoulli($p_i$).

$$
\log \left ( \frac{p_i}{1 - p_i} \right ) = s_0 + \sum_{j=1}^{p} s_j(x_{ij})
$$

## Additive hazard model

```{r}
expanded
```

```{r, echo=TRUE, eval=FALSE}
default_model = gam(default ~ s(install_num) + s(x),
                    data=df,
                    family='binomial')

prepay_model = gam(prepay ~ s(install_num) + s(x),
                   data=df,
                   family='binomial')
```

## Survival functions

```{r, echo=TRUE, eval=FALSE}
score_function = function(x, loan_term, default_model, prepay_model) {
    # initialize data frame to be scored
    x = data.frame(x)
    x = x[rep(1, loan_term),]
    x$install_num = 1:loan_term
    rownames(x) = NULL
    
    # calculate complement of hazard function by installment
    def_scr = predict(default_model, newdata=x)
    prepay_scr = predict(prepay_model, newdata=x)
    def_scr = 1 / (exp(def_scr) + 1)
    prepay_scr = 1 / (exp(prepay_scr) + 1)
    
    # calculate default and prepayment curves
    surv_func = data.frame(default_curve = cumprod(def_scr),
                           prepay_curve = cumprod(prepay_scr))
    
    return(surv_func)
}
```

## Amortization schedule

We also need to know the remaining principal at each installment and so have to compute a loan's amortization schedule.  For this we have to determine the (implicit) interest rate of the loan.

For a loan with loan amount $A$, payment amount $c$ and loan term $n$, the interest rate $r$ is the solution to the equation

$$
\frac{A}{\sum_{i=1}^{n} (1 + r)^{-i}} - c = 0 .
$$

## Interest calculation

```{r, echo=TRUE, eval=TRUE}
library(nleqslv)

get_interest = function(loan_term, loan_amount, payment_amount) {
    nleqslv(x=1,
            fn=function(x) {
                loan_amount / sum((1 + x)^(-(1:loan_term))) - payment_amount
                })$x
}
```

## Amortization schedule

```{r, echo=TRUE, eval=TRUE}
amortize = function(loan_term, loan_amount, payment_amount) {
    # initialize amortization schedule data frame and calculate interest
    schedule = data.frame(payment = rep(payment_amount, loan_term),
                          principal_payment = rep(0, loan_term),
                          interest_payment = rep(0, loan_term),
                          remaining_principal = rep(0, loan_term))
    rate = get_interest(loan_term, loan_amount, payment_amount)
    interest = loan_amount * rate
    
    # iteratively populate and return schedule
    schedule[1, 'interest_payment'] = interest
    schedule[1, 'principal_payment'] = payment_amount - interest
    schedule[1, 'remaining_principal'] = loan_amount - schedule[1, 'principal_payment']
    
    for (i in 2:loan_term) {
        schedule[i, 'interest_payment'] =  schedule[i - 1, 'remaining_principal'] * rate
        schedule[i, 'principal_payment'] =  payment_amount - schedule[i, 'interest_payment']
        schedule[i, 'remaining_principal'] =  schedule[i - 1, 'remaining_principal'] - 
            schedule[i, 'principal_payment']
    }
    return(schedule)
}
```

## Amortization schedule

```{r, echo=TRUE}
amortize(10, 1000, 120)
```

## Putting it all together

```{r, echo=TRUE, eval=FALSE}
compute_cash_flow(x,
                  loan_term,
                  loan_amount,
                  payment_amount,
                  default_model,
                  prepay_model,
                  j=0)
```

This returns a list that contains the amortization schedule, estimated default and prepayment curves, expected cash flow by installment, as well as the total expected cash.  The extra parameter `j` allows for making updated projections.

## Updating the projections

Given that a loan is current at time $j$ then for $i > j$

$$
\text{E}(Y_i) = c \frac{P(T_d > i)}{P(T_d > j)} \frac{P(T_p > i)}{P(T_p > j)} + c_i \frac{P(T_d > i)}{P(T_d > j)} \frac{P(T_p = i)}{P(T_p > j)} .
$$

The additional cash expected can be found by summing over $i \in \{j+1, \ldots, n \}$.

## References

- Hastie, T., Tibshirani, R.  "Generalized Additive Models." *Statistical Science*. Vol. 1, No. 3, 1986, pg. 297-318.

- Allison, P. "Discrete-Time Methods for the Analysis of Event Histories."  *Sociological Methodology*. Vol. 13, 1982, pg. 61-98.
